# Microarray analysis

In the sections below you'll find some reference implementations of several functions which will be of use to you in analyzing microarray datasets.


#  Loading multiple files

In the implementation below, you'll see an example function which loads microarray data from multiple input files.

``` {r}
loadTables <- function(dataPath) {
  
  #Load each .txt file in the path, parse the table with read.delim
  tables <- list()
  for (fileName in list.files(dataPath, pattern = ".txt")) {
    tables[[fileName]] <- read.delim(paste(dataPath, fileName, sep = "/"))
  }
  
  return(tables)
  
}
````

# Preprocessing

Here's a reference implementation of a preprocessing algorithm that performs many of the manipulations suggested in last week's lab:

``` {r}
preprocess <- function(table, dataName) {
  
  #Calculate Cy5 / Cy3 ratio (Ch1 / Ch2)
  ratios <- table$CH1D_MEAN / table$CH2D_MEAN
  
  #Filter bad (negative) ratio values
  ratios[ratios < 0] <- NA
  
  #Log2 transform ratios
  logRatio <- log(x = ratios, base = 2)
  logRatio[logRatio == NaN | logRatio == Inf | logRatio == -Inf] <- NA
  
  #Global median normalization (median Log2(Cy5/Cy3) == 0)
  medianRatio <- median(logRatio, na.rm = TRUE)
  normalizedRatios <- logRatio - medianRatio
  
  return(normalizedRatios)

}
````

# Hierarchical clustering

This wrapper function uses loadTables and preprocessing above to load a set of microarray tables from the DeRisi '96 data set and constructs a hierarchical cluster heatmap.:

``` {r}
analyzeArrays <- function(dataPath, platformFile) {
  
  #Load data.frame's from the given path
  tables <- loadTables(dataPath)
  
  #Load platform annotation information
  platform <- read.delim(platformFile)
  
  #Create a data.frame to hold the normalized data
  normalizedData <- data.frame(ORF = platform$ID..ORF)
  
  #For each file that was loaded as a data.frame in our list
  for (dataName in names(tables)) {
    
    message("Analyzing: ", dataName)
    
    table <- tables[[dataName]]
    normalizedValues <- preprocess(table, dataName)
    normalizedData[[dataName]] <- normalizedValues
    
  }
  
  #Create a matrix that we can use as the source for clustering with heatmap
  clusterMatrix <- as.matrix( normalizedData[2:ncol(normalizedData)] )
  
  #For simplicity, filter out any rows with missing data
  clusterMatrix <- clusterMatrix[apply(clusterMatrix, 
                                       MARGIN = 1, 
                                       FUN    = function(x) { 
                                                  !NA %in% x 
                                                }
                                       ),
                                 ]
  
  
  #Load gplots, which provides the colorpanel function
  library(gplots)
  
  #Use heatmap to perform clustering
  heatmap(clusterMatrix, 
          
          #Skip clustering columns
          Colv  = NA, 
          
          #Load a pretty Blue-to-black-to-yellow color pallet
          col   = colorpanel(100,"royalblue3","black","yellow"),
          
          #Set color saturation values at log2 of -2 to 2
          #zlim = c( -2, 2),
          
          #Hide gene labels because they will be too small to read
          labRow=c("")
          
          )
  
  return(normalizedData)
  
}
````

# Heatmaps of background intensity values

The function below takes a vector of background intensity values from the DeRisi '96 microarray dataset and makes a heatmap plot of those intesities across the surface of the microarray.

``` {r}
plotBackground <- function(backgroundIntensities, plotName = "") {

  sector1 <- matrix(backgroundIntensities[1:1600], nrow = 40, ncol= 40, byrow = TRUE)
  sector2 <- matrix(backgroundIntensities[1601:3200], nrow = 40, ncol= 40, byrow = TRUE)
  topBlocks <- cbind(sector1, sector2)
  
  sector3 <- matrix(backgroundIntensities[3201:4800], nrow = 40, ncol= 40, byrow = TRUE)
  sector4 <- matrix(backgroundIntensities[4801:6400], nrow = 40, ncol= 40, byrow = TRUE)
  bottomBlocks <- cbind(sector3, sector4)
  
  imageBlocks <- rbind(topBlocks, bottomBlocks)
  
  filled.contour(1:80, 1:80, 
                 imageBlocks, 
                 col  = heat.colors(20), 
                 main = paste("Background intensities (", plotName, ")", sep = "")
                 )
  
}
````